---
title: "SwiftKey NLP Capstone - Milestone Report"
author: "Farhan Bin Hossain"
date: "2026-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidytext)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Overview
This report presents a analysis of the SwiftKey Dataset provided by Johns Hopkins and Coursera.

---

## 1. Data Summary

```{r load-data, cache=TRUE}
setwd("/Users/keo/Documents/projects/swiftkey/dataset")

blogs   <- readLines("en_US/en_US.blogs.txt",   encoding = "UTF-8", skipNul = TRUE)
news    <- readLines("en_US/en_US.news.txt",    encoding = "UTF-8", skipNul = TRUE)
twitter <- readLines("en_US/en_US.twitter.txt", encoding = "UTF-8", skipNul = TRUE)

blogs_size   <- file.info("en_US/en_US.blogs.txt")$size   / 1024^2
news_size    <- file.info("en_US/en_US.news.txt")$size    / 1024^2
twitter_size <- file.info("en_US/en_US.twitter.txt")$size / 1024^2

summary_table <- data.frame(
  Source   = c("Blogs", "News", "Twitter"),
  Lines    = c(length(blogs), length(news), length(twitter)),
  Size_MB  = round(c(blogs_size, news_size, twitter_size), 1)
)

knitr::kable(summary_table, caption = "Dataset Summary")
```

The dataset is large — over **4 million lines** and **556 MB** of raw text. 
To keep analysis efficient, a random sample of 10,000 lines per source 
was used throughout this report.

---

## 2. Average Words Per Line

```{r sampling, cache=TRUE}
set.seed(42)
sample_blogs   <- sample(blogs,   10000)
sample_news    <- sample(news,    10000)
sample_twitter <- sample(twitter, 10000)
all_text <- c(sample_blogs, sample_news, sample_twitter)

source_words <- data.frame(
  Source = c("Blogs", "News", "Twitter"),
  Avg_Words_Per_Line = c(
    mean(sapply(strsplit(sample_blogs,   " "), length)),
    mean(sapply(strsplit(sample_news,    " "), length)),
    mean(sapply(strsplit(sample_twitter, " "), length))
  )
)

ggplot(source_words, aes(x = Source, y = Avg_Words_Per_Line, fill = Source)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#2166ac", "#4dac26", "#d01c8b")) +
  labs(title = "Average Words Per Line by Source",
       x = "Source", y = "Avg Words") +
  theme_minimal()
```

Blogs contain the longest text averaging around **42 words per line**, 
while Twitter is shortest at around **12 words** due to character limits. 
This difference will influence how the prediction model handles short 
vs long context.

---

## 3. Most Common Words

```{r wordcount}
text_df <- data.frame(text = all_text, stringsAsFactors = FALSE)

word_counts <- text_df %>%
  unnest_tokens(word, text) %>%
  anti_join(stop_words) %>%
  count(word, sort = TRUE) %>%
  filter(grepl("^[a-z]+$", word))

word_counts %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(word, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#6baed6", high = "#08306b") +
  labs(title = "Top 20 Most Common Words",
       subtitle = "Stopwords removed",
       x = "Word", y = "Count") +
  theme_minimal()
```

The most frequent words — **time, people, day, love, life** — reflect the 
personal and conversational nature of the data sources.

---

## 4. Bigrams (Two-Word Pairs)

```{r bigrams}
bigrams_filtered <- data.frame(text = all_text, stringsAsFactors = FALSE) %>%
  unnest_tokens(bigram, text, token = "ngrams", n = 2) %>%
  filter(!is.na(bigram)) %>%
  separate(bigram, c("word1", "word2"), sep = " ") %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word) %>%
  filter(grepl("^[a-z]+$", word1)) %>%
  filter(grepl("^[a-z]+$", word2))

bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ") %>%
  count(bigram, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(bigram, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#74c476", high = "#00441b") +
  labs(title = "Top 20 Most Common Bigrams",
       subtitle = "Two-word pairs (stopwords removed)",
       x = "Bigram", y = "Count") +
  theme_minimal()
```

---

## 5. Trigrams (Three-Word Sequences)

```{r trigrams}
data.frame(text = all_text, stringsAsFactors = FALSE) %>%
  unnest_tokens(trigram, text, token = "ngrams", n = 3) %>%
  filter(!is.na(trigram)) %>%
  count(trigram, sort = TRUE) %>%
  top_n(20, n) %>%
  ggplot(aes(x = reorder(trigram, n), y = n, fill = n)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#fdae6b", high = "#7f2704") +
  labs(title = "Top 20 Most Common Trigrams",
       subtitle = "Three-word sequences",
       x = "Trigram", y = "Count") +
  theme_minimal()
```

Trigrams like **"one of the"** and **"a lot of"** reveal natural English 
phrasing patterns that the prediction model will learn from.

---

## 6. Plan for Prediction Algorithm

The next word prediction model will work like:

1. **N-gram lookup** — When a usir types a word or phrase, the system will compare it with bigrams and trigrams and predict the next word.
2. **Backoff strategy** — If no trigram found, it will go back to bigram then unigram (most common word)
3. **Training** — For better acuracy the model need to be trained on larger dataset.
4. **Cleaning** — filtering and punctuation removal will be applied

---

## 7. Plan for Shiny App

The Shiny app will:

- Provide a **text input box** where users type a word or phrase
- Instantly display the **top 3 predicted next words**
- Show a simple bar chart of prediction confidence
- Be fast and lightweight by using a pre-computed lookup table

---

## Conclusion

The  analysis confirms the data is rich, diverse, and suitable 
for building a next word prediction. findings include significant 
differences in text length across sources, and clear patterns in word and 
phrase frequency that will directly inform the prediction algorithm.

